---
title: "ECON 474: Econometrics of Policy Evaluation"
subtitle: "Homework 3"
#author: "Marcelino Guerra (mguerra3)"
date: 'Due date: 11/11/2021'
#abstract: Requirements
output:
 bookdown::html_document2:  
  toc: true # table of content true
  toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
  #number_sections: true  ## if you want number sections at each table header
  theme: united  # many options for theme, this one is my favorite.
  highlight: tango  # specifies the syntax highlighting style
  fontsize: 16pt
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>


```{r setup, include=FALSE}
  knitr::opts_chunk$set(include = TRUE)  # TRUE for solution; FALSE for questions set
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(message = FALSE)
  knitr::opts_chunk$set(warning = FALSE)
  knitr::opts_chunk$set(fig.height = 5, fig.width = 7, out.width = '100%', fig.align = "center")
  options(width = 90)

```

```{css, echo=FALSE}
.solution {
background-color: #CCDDFF;
}
```


# Minimum Wage and Employment [50 points]

In this question, you will work with data from [Card and Krueger (1994)](https://davidcard.berkeley.edu/papers/njmin-aer.pdf). David Card and Alan Krueger collected information about fast-food restaurants in **New Jersey (NA)** and **eastern Pennsylvania (PA)** during two rounds of interviews. The first interview wave was in March 1992, just before New Jersey raised its minimum wage from 4.25 to 5.05 dollars - the raise occurred in April 1992. The second round of interviews was in November and December of 1992, 7 months after the new minimum wage policy in New Jersey. During that period, the minimum wage in Pennsylvania remained at the federal level of $4.25. The authors evaluated the impact of the new minimum wage law by analyzing this data. 

```{r, echo=F, message=F,warning=F, comment=F, out.width = '100%', align = "center"}
library(tigris)
library(sf)
library(tidyverse)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
states<- states()
states<-states%>%filter(STUSPS=="PA"|STUSPS=="NJ")
library(tmap)

tmap_mode("view")
tm_shape(states)+ 
  tm_polygons("NAME", palette="viridis", alpha=.5,legend.show = F,id="NAME", popup.vars=c("State " = "NAME")) +
  tm_text("NAME", size = 1.5)+
  tm_borders(lwd = 5, col = "black")+
  tm_basemap(server="OpenStreetMap",alpha=0.5)
```

Some columns have the same name but end with "2". That means those values correspond to the second round of interviews (after the minimum wage raise). 

```{r, echo=F}
library(kableExtra)

Variable<-c("sheet", "chain", "co_owned", "state", "wage_st", "wage_st2", "fte", "fte2" )
Definition<-c("Restaurant id", "1 if Burger King, 2 if KFC, 3 if Roy Rogers, and 4 if Wendy's", "1 if restaurant is company-owned", "1 if New Jersey, 0 if Pennsylvania" , "Starting wage (Dollar/hour) in March", "Starting wage (Dollar/hour) in Nov/Dec", "Full time equivalent employment in March", "Full time equivalent employment in Nov/Dec" )
df<-data.frame(Variable, Definition)
kbl(df, digits=2, caption = "Variables Description", booktabs = T) %>%
kable_styling(latex_options = c("striped", "HOLD_position"))%>%
  column_spec(2, width = "12cm")

```

Download the dataset [here](https://github.com/guerramarcelino/PolicyEval/raw/main/Datasets/ck94.RDS) and answer the following:

a. How many Burger King, KFC, Roy Rogers, Wendy's, and company-owned stores were located in NJ and PA? 

b. Calculate four averages: the average **starting wages** (`wage_st`) in NJ and PA before and after the new minimum wage law in NJ. What is the difference-in-differences estimate of the impact of minimum wage on fast food starting wage in NJ?

c. Repeat the same exercise as in (b) for **full-time equivalent employment** (`fte`). What is the impact of the new minimum wage law on employment in NJ fast-food restaurants?

d. Download the panel data version of this dataset [here](https://github.com/guerramarcelino/PolicyEval/raw/main/Datasets/ck94_panel.RDS). Then, run the following regression DiD model 

$$Y_{ist}=\alpha+\beta Dint_{st} + \lambda POST_{t} + \gamma TREAT_{s}+\varepsilon_{st}$$
using wages and full time equivalent employment. How do those estimates compare to the results in (b) and (c)?

*Hint:* you need to create the variables *Dint*, *POST*, and *TREAT*. 

e. A nice feature of regression DiD is that you can control for other factors. For instance, you might want to add covariates such as chains (`chain`) and a dummy to capture whether the restaurant is company-owned (`co_owned`). Repeat the models you used in (d), adding those covariates. Do your results change too much when adding restaurant-specific covariates? Do you have an explanation for that? 


f. An alternative to comparing NJ and PA restaurants is to use restaurants within NJ with high and low wages before the minimum wage increase. Restrict the sample to NJ and identify restaurants paying salaries above and below 5 dollars/hour (i.e., create a dummy that takes on 1 if the restaurant pays salary below 5 dollars). Then, compare employment and wages before and after the new minimum wage law between restaurants above and below the $5 threshold. What is the relative impact of the minimum wage on employment within NJ? How do the within NJ estimates compare to those obtained in part (d)?

*Hint:* Use the first dataset for (f) and (g)

g. Now, do the same as in (f) for PA. What is your difference-in-difference estimate when faking the treatment for PA? Why is this considered a robustness check for the DiD strategy in this setting?



# Do Police Reduce Crime? [50 points]

Common sense dictates that more police on the streets reduce criminal activity. It is also a prediction from the standard model of the economics of crime, whenever that increase in policing reflects a higher probability of apprehension perceived by offenders [(Becker 1968)](https://www.nber.org/system/files/chapters/c3625/c3625.pdf). Nevertheless, to establish a causal effect between those two, you need to break the circle: more crime leads to more police on the streets. One way to do it is through natural experiments. 

[Di Tella and Schargrodsky (2004)](https://www.aeaweb.org/articles?id=10.1257/000282804322970733) exploited the variation in policing levels induced by terrorist acts. In 1994, there was a terrorist attack on a Jewish center in Buenos Aires, Argentina. That act increased the intensity of policing in blocks of Buenos Aires that had either a Jewish or Muslim institution for the following months. Presuming that the spatial distribution of those centers is exogenous in a police-crime regression, it is possible to estimate the causal effect of police on crime in this setting. Di Tella and Schargrodsky (2004) define blocks with a Jewish institution as a treatment group while considering areas farther away the control group. The outcome of interest is car theft.   

```{r, echo=F}
library(kableExtra)

Variable<-c("observ", "barrio", "calle", "altura", "institu1", "institu3", "distanci", "edpub",   
 "estserv","banco","totrob","mes" )
Definition<-c("Block id", "Neighborhood", "Street", "Street Numbering", "1 if there is a Jewish institution in the block, 0 otherwise", "1 if there is a Jewish institution one block away, 0 otherwise", "Distance to closest Jewish institution (in blocks)", "1 if there is a public building/embassy, 0 otherwise", "1 if there is a gas station, 0 otherwise", "1 if there is a bank, 0 otherwise", "Car Theft", "Month")
df<-data.frame(Variable, Definition)
kbl(df, digits=2, caption = "Variables Description", booktabs = T) %>%
kable_styling(latex_options = c("striped", "HOLD_position"))%>%
  column_spec(2, width = "12cm")

```

Download the data [here](https://github.com/guerramarcelino/PolicyEval/raw/main/Datasets/DS04.RDS) and answer the following:

a. Run a `t.test()` to compare the average car theft on the period April-July (`months==4, 5, 6, and 7`) in blocks with a Jewish institution (`institu1==1`) and in blocks farther away (`institu1==0`). Is there any difference in outcomes during the pre-treatment period? 

b. Run a `t.test()` to compare the average car theft on the period August-December (`months==8, 9, 10, 11 and 12`) in blocks with a Jewish institution (`institu1==1`) and in blocks farther away (`institu1==0`). Is there any difference in outcomes during the treatment period? 

c. What is the difference-in-differences estimate of the impact of more police on car theft? 

d. Plot the evolution of average car theft in the treated (`institu1==1`) and the control (`institu1==0`) city blocks from April to December. What can you say about the common trends assumption in this setting?

*Hint:* average out the outcomes using `group_by` and `summarize()`. You want to `group_by()` treatment status and also month (`mes`).

e. Estimate the following model (clustering standard errors at the block level):

$$\text{Car Theft}_{it}= \beta \text{Same Block Police}_{it}+ Month_{t}+Block_{i}+ \varepsilon_{it}$$
where `Same Block Police` is the interaction between `institu1` (treated blocks) and the treatment period (from August to December).

f. To interpret the results, get the average number of car thefts from August to December in the control group - you already have that number from (c). Then, divide your DiD estimate by that average and multiply by 100. What is the percentage reduction in car theft due to the increase in policing?

g. Considering a roughly approximated increase in police presence of 223%, what is the crime-police elasticity in this setting? 

*Hint:* divide the reduction in car theft by the increase in police presence



