---
title: "Synthetic Control"
author: "Marcelino Guerra"
date: "Last update: 11/03/2021"
abstract: 
output: 
  rmdformats::readthedown
extra_dependencies: ["cancel", "amsmath", "amssymb", "float","tcolorbox","awesomebox"]
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
  knitr::opts_chunk$set(include = TRUE)  # TRUE for solution; FALSE for questions set
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(message = FALSE)
  knitr::opts_chunk$set(warning = FALSE)
  knitr::opts_chunk$set(fig.height = 6, fig.width = 11, out.width = '100%', fig.align = "center")
  options(width = 90)
library(fontawesome)
```


## Preliminaries

Here we replicate some results from [Abadie, Diamond and Hainmueller (2015)](https://economics.mit.edu/files/11858). Results are similar but not the same since I ignore the cross-validation exercise to choose the *v weights*. Nevertheless, conclusions remain the same. Original replication files can be found [here](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/24714). 

Download the dataset `germany.RDS` [here](https://github.com/guerramarcelino/PolicyEval/raw/main/Datasets/germany.RDS), and check lecture notes for discussion [here](https://guerramarcelino.github.io/Econ474/Lectures/Lec7/lec7?panelset4=synthetic-control-results2&panelset5=gap-in-homicide-rates2&panelset6=synthetic-control-weights2&panelset7=balance-table2#19). 

## The Economic Cost of the 1990 German Reunification


```{r}
data<-readRDS("germany.RDS")
```


```{r, message=F, warning=F, results=F}
library(tidyverse)
library(Synth)
dataprep.out <-
  dataprep(foo = data,
           predictors    = c("gdp","trade","infrate"),
           predictors.op = "mean",
           time.predictors.prior = 1960:1990,
           dependent     = "gdp",
           unit.variable = "index",
           unit.names.variable="country",
           special.predictors = list(
           list("industry" ,1981:1990, c("mean")),
           list("schooling",c(1980,1985), c("mean")),
           list("invest80" ,1980, c("mean"))),
           time.variable = "year",
           treatment.identifier = 7,
           controls.identifier = c(1,2,3,4,5,6,8,9,10,12,14,16,18,19,20,21),
           time.optimize.ssr = 1960:1990,
           time.plot = 1960:2003
         )
```

```{r, results=F,}
synth.out <- synth(dataprep.out)
```

```{r}
synth.tables <- synth.tab(dataprep.res = dataprep.out,
                          synth.res = synth.out)

synth.tables
```

```{r}
library(kableExtra)

weights<-data.frame(synth.tables$tab.w)
weights<-weights%>%select(-unit.numbers)
names(weights)<-c("Weight", "Country")
kbl(weights, digits=2, caption = "Weights Table", booktabs = T) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r}
balance<-synth.tables$tab.pred
rownames(balance)<-c("GDP per capita", "Trade openness", "Inflation rate", "Industry share", "Schooling", "Investment rate")
kbl(balance, digits=2, caption = "Balance Table 1960-2003", booktabs = T) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```


```{r}
path.plot(synth.res = synth.out, dataprep.res = dataprep.out,
          Ylab = "Per capita GDP (PPP, 2002 USD)", Xlab = "Year",
          Ylim = c(0, 35000), Legend = c("West Germany","Synthetic West Germany"), Legend.position = "bottomright",
          tr.intake = 1990)
```


```{r}
gaps.plot(synth.res = synth.out, dataprep.res = dataprep.out,
           Ylab = "Per capita GDP (PPP, 2002 USD)", Xlab = "Year",
          tr.intake = 1990)

```


## Placebo Studies


```{r, warning=F, message=F, results=F}
dataprep.out2 <-
  dataprep(foo = data,
           predictors    = c("gdp","trade","infrate"),
           predictors.op = "mean",
           time.predictors.prior = 1960:1975,
           dependent     = "gdp",
           unit.variable = "index",
           unit.names.variable="country",
           special.predictors = list(
           list("industry" ,1971:1975, c("mean")),
           list("schooling",c(1970,1975), c("mean")),
           list("invest80" ,1980, c("mean"))),
           time.variable = "year",
           treatment.identifier = 7,
           controls.identifier = c(1,2,3,4,5,6,8,9,10,12,14,16,18,19,20,21),
           time.optimize.ssr = 1960:1975,
           time.plot = 1960:1990
         )
synth.out2 <- synth(dataprep.out2)
path.plot(synth.res = synth.out2, dataprep.res = dataprep.out2,
          Ylab = "Per capita GDP (PPP, 2002 USD)", Xlab = "Year",
          Ylim = c(0, 35000), Legend = c("West Germany","Synthetic West Germany"), Legend.position = "bottomright",
          tr.intake = 1975)
```


```{r, message=F, warning=F, results=F}
library(SCtools)
placebo <- generate.placebos(dataprep.out = dataprep.out,
                             synth.out = synth.out, strategy = "multiprocess")
```

```{r}
plot_placebos(placebo)
```

```{r}
test_out <- mspe.test(placebo)
test_out$p.val
```

```{r}
mspe.plot(tdf = placebo)
```